\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{parskip}
\usepackage{xurl}

\author{Andrea Riolo Vinciguerra}
\title{Compilation Techniques: Laboratory Project\\ Project Report}
\date{\today}

\newcommand{\Code}[1]{\texttt{#1}}

\begin{document}

\maketitle
\clearpage
\tableofcontents
\clearpage

\section{Introduction}

This project is organized in a single codebase that houses two compilers:
MiniImp and MiniFun. Both share common background code (like the diagnostics
system) and the core logic is kept strictly separate from the main user
interface (via command line). This setup allows users to easily choose between
generating a control flow graph, running an interactive interpreter, or
executing a code file.

A main focus of the design is modularity. By keeping the code well-organized and
adaptable, the system is much easier to maintain and expand over time. Comments
are included (more or less helpful) throughout the source files to make
comprehension easier.

The code is organized by separating distinct functional blocks into different
files. For example, the core evaluator (or interpreter) and the Abstract Syntax
Tree definitions for both compilers are stored in a file called
\path{engine.ts}. The rest of the functionality builds upon this file. For
example, MiniImp extends this core engine to construct its control flow graph
and generate an intermediate representation, while MiniFun supports a complete
polymorphic type system.

I chose TypeScript as the programming language for two main reasons. First, I
wanted the opportunity to master a language I was less familiar with. Secondly,
TypeScript is excellent at handling union types. This feature (used extensively
throughout the program) makes writing compiler code much simpler by avoiding the
bulky, complex structures often needed in languages like C++ or Rust.

Because Node.js works across different operating systems, the project can be
packaged into a single executable for Linux, macOS and Windows. On the latter,
however, the user must run the program manually by invoking the \texttt{node}
command.

\section{Running the code}

The software can be run in two ways: automatically or manually. The automatic
method is designed specifically for Linux or macOS. Instead, the manual approach
works on any operating system.

Either approach requires that Node.js and its package manager, \texttt{npm}, are
installed and properly configured in the system's \texttt{PATH}.

\subsection{Automatic run}

Inside the main project folder, there is a helpful script called
\path{run.sh}. This script handles the setup automatically by downloading the
required files, building the chosen compiler (MiniImp or MiniFun), and launching
the program.

Running either compiler is straightforward. For example, to start MiniImp, this
command is sufficient:
\begin{verbatim}
$ ./run.sh miniimp --help
\end{verbatim}

Similarly, for MiniFun:
\begin{verbatim}
$ ./run.sh minifun --help
\end{verbatim}

Any extra instructions or arguments should be typed directly after the
compiler's name. As shown in the examples above, adding the \texttt{--help} flag
is a great way to see a full list of all the available commands.

\subsection{Manual compilation}

Manual compilation is available. In such case, these commands must be executed
upon downloading the repository:
\begin{verbatim}
$ npm ci
$ npm run build:miniimp     # or npm run build:minifun
$ dist/miniimp --help       # or dist/minifun --help
\end{verbatim}

For Windows, the last step involves invoking \texttt{node} explicitly, like so:
\begin{verbatim}
$ node dist/miniimp --help  # or node dist/minifun --help
\end{verbatim}

From now on, the compiler program will be referenced as \texttt{./run.sh
<compiler>} for consistency.

\subsection{Generating MiniImp executables}

MiniImp's frontend generates basic LLVM intermediate representation, which
cannot be executed directly. To create a runnable program, this IR must be
processed by an assembler and a linker.

Before converting the intermediate representation into machine code, it must be
optimized. This step removes inefficient patterns, such as the \texttt{alloca}
trick, which is a technique used for simplifying the generation of LLVM IR. This
is explained in detail in a dedicated section.

To make the compilation process easier for Linux and macOS users, a
\path{compile.sh} script is provided. For example, to compile a MiniImp file
named \path{code.mi} into an executable, the following command is used:
\begin{verbatim}
$ ./compile.sh code.mi
\end{verbatim}

The intermediate temporary files are deleted by default, but can be preserved by
supplying the argument \texttt{-k} before the filename, like so:
\begin{verbatim}
$ ./compile.sh -k code.mi
\end{verbatim}

For manual compilation, standard LLVM tools can be used to generate the final
\path{miniimp} binary:
\begin{verbatim}
$ ./run.sh miniimp -f code.mi -c >out.ll
$ opt -p=mem2reg out.ll -S -o opt.ll
$ llc -filetype=obj opt.ll -o miniimp.o
$ clang misc/wrapper.c miniimp.o -o miniimp
\end{verbatim}

As an additional note, I modified the C wrapper slightly to allow for the
arguments to be passed via command line, rather than standard input.

\section{Diagnostics system}

Both compilers share the same diagnostics system, ensuring consistent and
precise error reporting for the user. The interpreters intentionally bypass this
system, as runtime evaluation errors represent a failure in code validation
rather than a user error.

To provide accurate diagnostics, I made Abstract Syntax Tree nodes track
their exact location in the source file using a \Code{SourceSpan}. This
structure records the starting and ending lines and columns without needing to
store a copy of the code. When a \Code{DiagnosticError} is thrown, the caller
invokes its \Code{format} function and passes in the full source code. Using the
\Code{SourceSpan}, this function derives a detailed, colored error message that
highlights the exact problem area, along with a clear explanation and any
helpful additional notes.

As an example, here is the error presented to the user in case they used an
undefined variable in their code:
\begin{verbatim}
Error: undefined variable 'test'
11 |         c := test + 5;
                  ^^^^
\end{verbatim}

%TODO: anyone else?
The implementation is located in the file \path{src/diag.ts} and the
functionality is used by all the code validators and parsers.

\section{MiniFun: the functional language}

MiniFun is a functional language supporting three types of first-class values:
functions, integers and booleans. It supports let bindings and recursion.

\subsection{Abstract Syntax Tree and evaluation}

The type definitions for the Abstract Syntax Tree can be found in
\path{src/minifun/engine.ts}. In this file, the type \Code{Value} represents
the final result of an evaluation. There are three possible values:
\Code{number}, \Code{boolean} and \Code{Closure}. The first two rely
on TypeScript's built-in types, while the last one is a custom structure. A
basic expression is represented through the type \Code{Expr}, which is a
discriminated union of all the available operators.

A closure is a special function that ``remembers'' its surrounding environment
at the time of definition and uses it during execution. Because of this, its
type definition must include both the function's body and the saved environment.

As seen in the grammar of the language, some expressions like \Code{let} or
\Code{fun} can accept type labels to override automatic type inference. I
made these labels optional, giving users the choice to either specify types
manually or rely on (and trust) the compiler's type system.

The interpreter processes the Abstract Syntax Tree generated by the parser. It
traverses the tree recursively, evaluating sub-nodes, updating the environment
and returning the final value. This is what the \Code{evalExpr()} function
does.

\subsubsection{Environment}

Unlike imperative languages, purely functional languages do not allow side
effects. Instead of modifying global memory, an evaluated expression returns an
updated environment. As the interpreter reaches the end of the tree and climbs
back up the recursive chain, these temporary inner environments are discarded
until the original (empty) one is reached.

To manage this, I used an \Code{Environment} interface with two main functions:
one to look up the value of a specific variable, and another to create a fresh
copy of the same environment with a newly assigned identifier. This interface is
implemented trivially using a linked list, but is left purposefully generic to
allow for more solutions.

\subsubsection{Inference rules}

The evaluation inference rules are implemented exactly as they are shown in the
slides, with the exception of recursive function calls (declared with
\texttt{letfun}).

Instead of using a special rule to apply recursive functions, which injects the
function's name into the environment before evaluating its body, I modified the
closure's saved environment to point directly back to the closure object itself.
This circular reference allows the recursive function to be called just like any
other function, since it is already available within its own stored environment.

The new closure definition rule, therefore, becomes as follows:
$$
\text{LetFun}\quad
\frac
{ \rho' = \rho[f \mapsto (f, x, t_1, \rho')] \vdash t_2 \longrightarrow t }
{ \rho \vdash \texttt{letfun} \; f \; x \; = \; t_1 \; \texttt{in} \; t_2
\longrightarrow t }
$$
Notice how the function's identifier no longer points to the original
environment $\rho$, but rather to the modified environment $\rho'$.

%$$
%\text{FunApp}\quad
%\frac
%{ \rho \vdash t_1 \longrightarrow (x, t_1', \rho') \quad \rho \vdash t_2
%\longrightarrow t_2' \quad \rho'[x \mapsto t_2'] \vdash t_1' \longrightarrow t }
%{ \rho \vdash t_1t_2 \longrightarrow t}
%$$

\subsection{Type system}

%TODO
The type definitions and implementation for MiniFun's type system are located in
\path{src/minifun/validator.ts}. I implemented the complete polymorphic
type system in the slides, which is based on the Hindley-Milner algorithm. The
whole system revolves around the concept of \textit{monotypes} and
\textit{polytypes}, each defined in its own structure.

A monotype either represents a simple concrete type, like integer or boolean, or
a generic, undefined type. In such case, the type is represented by a variable.
If no extra context is provided, the type stands for itself. An example would be
the identity function (\Code{fun x => x}), which has type \Code{a -> a}.

A polytype, on the other hand, is essentially a template waiting to be assigned
a more specific type. In the code, this is handled by wrapping a standard
monotype inside a structure that keeps track of its bound, yet unassigned
variables. This setup allows the system to generate various monotypes from a
single polytype, making it possible to reuse the same expression safely in
different contexts. As for the example above, the identity function may be
defined once and take the types \Code{int -> int} or \Code{bool -> bool} as
needed.

\subsubsection{Context}

Structurally, the type system's context is very similar to the evaluator's
environment. However, I modified it to support an iterator operation that
can go through all currently defined identifiers. This makes it easier to apply
substitutions to the entire context.

Once again, the interface is separated from the actual implementation to favor
modularity and code reuse.

\subsubsection{Substitutions}

A substitution works by mapping a type variable to a specific type or to another
variable. In the latter case, it enforces that both generic types must match.
For example, a substitution could look like \Code{\{a -> int\}} or \Code{\{b ->
a\}}.

Different sets can be composed. For instance, to calculate the composition
$S2S1$, we must first apply $S2$ to all the monotypes present in $S1$, and then
include any remaining substitution from $S2$. The recursive
\Code{composeSubst()} function is designed to do exactly that.

A single substitution pass can update multiple variables at the same time
because these mappings are transitive. If multiple variables are linked, they
will all resolve to the same final type. For example, given the substitution
\Code{\{a -> int, b -> a, c -> b\}}, the variables \Code{a}, \Code{b} and
\Code{c} will all evaluate to \Code{int}.

Because of this transitive behavior, a variable cannot point to itself or a type
that contains it. To prevent infinite loops during substitution application, the
\Code{bindVar()} function, which handles the assignment of a single type
variable, first checks for these circular references.

Substitutions can also be applied to polytypes. Since a polytype is a template
that binds certain variables from its inner monotype, the type system only needs
to apply substitutions to the remaining free variables. The
\Code{freeVarsPoly()} function is responsible for identifying exactly which
variables are unbound and available for substitution.

Finally, substitutions can be applied on a whole context. This is done by
iterating through the list of all defined identifiers and updating only the
relevant (and unbound) type variables.

\subsubsection{Algorithm W}

The implementation of Algorithm W's inference rules follows the slides. However,
I extended these rules to support optional type annotations on \texttt{fun} and
\texttt{letfun} statements. When a type label is omitted, the compiler relies on
the standard inference rules. If a label is provided, the system enforces
specific type restrictions during the inference process.

Type transformation functions are implemented exactly as per slides:
\begin{itemize}
    \item \Code{unify()}: solves type constraints by finding a substitution that
        makes two types identical. It may fail and yield a diagnostic error.
    \item \Code{gener()}: converts a monotype into a polytype by binding free
        type variables that are not referenced in the current context.
    \item \Code{inst()}: replaces bound variables in a polytype with fresh type
        variables, allowing generic functions to be used with different concrete
        types and making polytypes ``act'' as true templates.
\end{itemize}

The inference rule for the type-labelled \texttt{fun} statement is relatively
straightforward. It simply assigns the supplied type to the argument rather than
generating a fresh variable:
$$
\text{TFun}\quad
\frac
{ \Gamma [x \mapsto \tau] \vdash t : \tau',S }
{ \Gamma \vdash \texttt{fun} \; x : \tau \; => \; t : S\tau \rightarrow
  \tau',S }
$$

The \texttt{letfun} statement requires that the type label be enforced during
unification instead, since we supply the whole function type:
$$
\text{TLetFun}\quad
\frac
{
    \begin{array}{c}
        \tau' \text{ fresh} \quad
        \tau'' \text{ fresh} \quad
        \Gamma [f \mapsto \tau'] [x \mapsto \tau''] \vdash t_1 : \tau_1,S_1 \\
        S_2 = \text{unify}(S_1\tau', S_1\tau'' \rightarrow \tau_1) \quad
        S_3 = \text{unify}(\alpha \rightarrow \beta, S_1\tau'' \rightarrow
            \tau_1) \\
        S_3S_2S_1\Gamma [f \mapsto S_3S_2S_1\tau'] \vdash t_2 : \tau_2,S_4
    \end{array}
}
{ \Gamma \vdash \texttt{letfun} \; f \; x : \alpha \rightarrow \beta \; = \;
  t_1 \; \texttt{in} \; t_2 : \tau_2,S_4S_3S_2S_1 }
$$

\section{MiniImp: the imperative language}

MiniImp is a simple imperative language that works by receiving an input value
through one variable and expecting the program to save its final result into a
different variable. The language supports assignments, conditionals, loops and
sequences.

\subsection{Abstract Syntax Tree and evaluation}

The definitions for the Abstract Syntax Tree are located in
\path{src/miniimp/engine.ts}. Unlike MiniFun, the primary building block here is
a command (represented by the \Code{Cmd} discriminated union). Because of this
imperative design, evaluating a command does not return a final value. Instead,
the program operates through side effects, storing its calculations directly in
memory. Numeric and boolean expressions are used exclusively within assignments
or as comparisons inside conditional statements.

The \Code{evalCmd()} executes commands step-by-step, updating the memory state
and evaluating any necessary sub-expressions.

\subsubsection{Memory}

The \Code{Memory} interface serves as the program's runtime environment. Instead
of creating a new copy of the environment for every step, the system uses a
single, shared instance throughout the entire evaluation process. This instance
provides two simple methods for managing data: \Code{update()} and
\Code{read()}. Trivially, this interface is simply implemented using a standard
TypeScript map.

\subsection{Control Flow Graph}

To analyze the correctness of MiniImp programs, we cannot rely on a standard
type-checking system. Because imperative side effects require evaluating all
potential branching paths, this problem is instead solved using a
\textit{control flow graph}.

All functions related to generating, exporting and manipulating these graphs are
located in the file \path{src/miniimp/graph.ts}.

\subsubsection{Minimal graph}

A minimal graph is represented by the \Code{Graph} structure, while individual
nodes are defined by the \Code{Node} structure. To further restrict operations
at the type-system level, derived utility types like \Code{ExitNode} are also
included.

There are three types of nodes: \Code{skip}, \Code{assign} and \Code{cond}. Each
of these is assigned a specific command from the Abstract Syntax Tree.

The \Code{genGraph()} function inductively generates this minimal graph,
ensuring that each block contains at most one instruction. A strict invariant is
maintained throughout this process: every sub-graph must have exactly one entry
node and one exit node, though these can be the same node.

The algorithm scans the AST and builds a suitable sub-graph for each command
while carefully preserving the invariant. Finally, these sub-graphs are linked
together using the \texttt{seq} pseudo-command.

To respect the invariant at all times, the generator may insert extra
\texttt{skip} commands not present in the user's original code. These are
referred to internally as "fake skips" and are represented as standard
\texttt{skip} nodes without an associated AST command.

\subsubsection{Maximal graph}

A maximal graph is represented by the \Code{BlockGraph} structure, with its
nodes described by the \Code{Block} structure. Rather than generating a maximal
graph from scratch, I created a helper function called \Code{maximizeGraph()}
that converts an existing, verified minimal graph into a maximal one.

The structural invariant remains similar, but with an important nuance: the
distinction between normal command blocks and ``merge'' blocks. A merge block
functions like a standard block, but specifically incorporates the first "fake
skip" following the branches of a conditional statement. I introduced this
primarily to keep debugging visualizations clean, as it prevents the graph from
generating an empty block when a program ends immediately after an \texttt{if}
statement. Although merge and standard blocks are functionally equivalent, this
specific design safely groups sequential commands while guaranteeing a single
exit block for branching statements.

\subsubsection{Minimal or maximal?}

When determining the ideal graph structure for the compiler, I opted to use the
minimal graph as the primary format because it significantly simplifies the
analysis and optimization phases. Since each block contains at most one
instruction, we can apply straightforward inductive algorithms without needing
special heuristics or complex inner-block analyses.

Furthermore, generating a minimal graph requires fewer steps and is more
structural, which is exceptionally good for recursive functions.

However, the graph maximizer remains essential for the code generation phase.
Grouping instructions into larger blocks allows the compiler to produce clean,
readable LLVM IR that is not cluttered with superfluous jump instructions.
Additionally, developing this conversion tool was an excellent exercise for my
fluency in TypeScript.

\subsubsection{Exporting graphs}

To assist with debugging and development, I created a generic graph-exporting
helper called \Code{graphToDOT()}. This function generates a DOT-formatted
string representing the control flow graph. The resulting output can be
visualized using online DOT viewers or the standard \texttt{dot} command-line
utility.

Rather than hardcoding the export logic, this routine accepts a set of callback
functions as arguments. Each function plays a specific role in defining the
structure and appearance of the exported graph.

These callback functions include:
\begin{itemize}
    \item \Code{labelShape(node)}: determines the DOT format label and visual
        shape for a given node.
    \item \Code{isBranching(node)}: identifies whether the node represents a
        branching point in the control flow.
    \item \Code{getNext(node)}: retrieves the subsequent node(s). This can
        return null at an exit, a single node for sequential commands, or an
        array of two nodes for conditional statements.
    \item \Code{skipElem(node)}: evaluates whether a node should be omitted from
        the visual output, which is particularly useful for hiding internal
        "fake skip" nodes.
\end{itemize}

Because of this modular design, \Code{graphToDOT()} can be easily adapted by
creating specific wrappers for different graph types. This flexibility allows
the compiler to export both minimal and maximal graphs, as well as seamlessly
attach extra analysis metadata to each node.

To quickly inspect basic graphs, the module provides two convenient wrappers:
\Code{exportGraph()} and \Code{exportBlockGraph()}. A DOT graph can be generated
for a specific program by executing the following commands:
\begin{verbatim}
$ ./run.sh miniimp -f code.mi -g     # minimal graph
$ ./run.sh miniimp -f code.mi -g -m  # maximal graph
\end{verbatim}

\subsection{Graph analysis}

With the generated control flow graph, the compiler must perform data flow
analysis to extract information about the execution paths. This process is
needed for both verifying code correctness, such as catching undefined
variables, and for enabling optimizations.

The logic for these operations is located in the file
\path{src/miniimp/analysis.ts}.

\subsubsection{The work queue algorithm}

Data flow analysis involves propagating information across the graph until a
stable fixpoint is reached. To handle this efficiently, I implemented a generic
\Code{wqAnalyze()} function.

This method uses a queue-based breadth-first search algorithm. It enqueues graph
nodes and applies a custom local update function to each one. If a node's
internal state changes during this update, its neighboring nodes (either
successors or predecessors, depending on the direction of the analysis) are
added back to the work queue.

The \Code{wqAnalyze()} generic wrapper reduced code duplication when
implementing multiple distinct analyses, and is driven by two primary callback
functions provided as arguments:
\begin{itemize}
    \item \Code{init(node)}: establishes the initial state for a single node
        before the main analysis begins. This function is executed once for
        every node in the graph, relying on the surrounding scope to set up the
        necessary data structures.
    \item \Code{local(node, ctx, wq, map)}: defines the core update logic
        applied to each node as it is processed from the work queue. It receives
        the current node, a context object (\Code{ctx}) containing its immediate
        neighbors, the current state of the analysis (\Code{map}), and a
        reference to the work queue (\Code{wq}). Providing direct acces to the
        queue allows the function to re-enqueue neighboring nodes, ensuring that
        state changes are propagated across the graph until a stable state is
        reached.
\end{itemize}

\subsubsection{Code specifics and types}

All the analyses work by assigning a special structure to each graph node via a
standard TypeScript map. For the \textit{Defined} and \textit{Live Variables}
analyses, the structure \Code{VarSets} is used (aliased to \Code{DefinedVars}
and \Code{LiveVars}, respectively), which is a pair of identifier sets: one for
the \textit{in} variables and one for the \textit{out} variables.

The \textit{Reaching Definitions} analysis, on the other hand, uses the
\Code{DefSets} structure, which stores the nodes where the assignment occurs by
reference. This avoids having to keep a separate map in memory to assign
definition nodes a unique identification number.

\subsubsection{Defined Variables and validation}

The function \Code{analyzeDefinedVars()} handles the most critical analysis, the
\textit{Defined Variables}. Using the output from this process,
\Code{checkUndefined()} scrubs the graph to ensure that no command ever attempts
to read a variable before it has been assigned a value. If an uninitialized
variable is detected, the validation process fails and the compiler throws a
\Code{DiagnosticError} with the precise location of the error in the source
code.

\subsubsection{Live Variables and Reaching Definitions}

The \textit{Live Variables} and \textit{Reaching Definitions} analyses are
handled by two other functions: \Code{analyzeLiveVars()} and
\Code{analyzeReaching()}. These match the specifications in the slide exactly.

\subsubsection{Graph visualization}

To aid in debugging and verify the correctness of the analyses' logic, the
module includes two wrappers of the generic \Code{graphToDOT()} graph exporting
helper.

\Code{exportDefinedVars()}, \Code{exportLiveVars()} and
\Code{exportReachingDefs()} convert the internal graph structures into DOT
format strings that display the calculated \textit{in} and \textit{out} data
sets for every single node in the control flow graph.

To generate a DOT graph for a specific analysis, these commands may be executed:
\begin{verbatim}
$ ./run.sh miniimp -f code.mi -g -ad  # defined variables
$ ./run.sh miniimp -f code.mi -g -al  # live variables
$ ./run.sh miniimp -f code.mi -g -ar  # reaching definitions
\end{verbatim}

\subsection{Optimization}

Once analyses have been performed on the control flow graph, the compiler enters
the optimization phase. The goal of this step is to produce a more efficient
graph by stripping away redundant instructions and pre-calculating predictable
values at compile time, before handing off the final result to the code
generation phase.

The optimization logic is housed in \path{src/miniimp/opt.ts} and consists of
three distinct passes that work together to simplify the graph.

\subsubsection{Dead Store Elimination}

\textit{Dead Store Elimination} is a structural optimization that removes
useless assignments from the code. The pass is implemented in the \Code{dse()}
function according to the slides, and relies on the \textit{Live Variables}
analysis.

The algorithm iterates through all the nodes in the control flow graph. If it
encounters an \Code{assign} node, it checks the \Code{liveOut} set for that
block. If the assigned variable does not appear in the \textit{out} set, it
means that the variable is never read by any subsequent instruction. Therefore,
the assignment is completely redundant and can be safely deleted.

Once again, the advantage of using a minimal graph is evident: because a graph
node can only contain one instruction, there's at most one assigned variable for
every single node.

Because removing a node structurally alters the graph, the predecessor and
successor links must be re-routed to avoid breaking the graph's invariants.
After a successful deletion, the graph is permanently changed, meaning the data
flow maps are now stale. Therefore, as seen in the code's comments, the DSE pass
requires that all analyses be re-run, and the pass repeats until no further dead
stores are found.

\subsubsection{Constant Folding}

\textit{Constant Folding} is a local optimization that simplifies expressions
before runtime. The \Code{constFold()} function traverses both numeric and
boolean expressions within the Abstract Syntax Tree.

When the compiler detects an operation involving only raw literal values, it
evaluates the result immediately and replaces the entire sub-tree with a single
value node.

The numeric folding logic incorporates arithmetic identities, such as stripping
away \Code{x + 0} or \Code{x * 1}, and converting \Code{x * 0} to \Code{0}. It
even performs simple associative folding, automatically reducing expressions
like \Code{10 + (20 + x)} directly down to \Code{30 + x}.

I thought of creating a more complex algebraic system to keep track of the
variable factors used in mathematical formulas, so as to simplify expressions
such as \Code{x + 3 * y + 5 * x + 2 * y} into \Code{6 * x + 5 * y}, but that
seemed slightly off the scope of the project.

Unlike DSE, Constant Folding only mutates the inner AST nodes and never alters
the structural layout of the graph itself.

\subsubsection{Constant Propagation}

\textit{Constant Propagation} complements Constant Folding by replacing
variables with their known literal values. It is implemented via
\Code{constProp()} and follows the slides exactly, leveraging the
\textit{Reaching Definitions} analysis.

Whenever an expression uses a variable, the compiler checks the \textit{in} set
of reaching definitions for that specific node. If there is exactly one
assignment reaching the node, and that assignment sets the variable to a raw
literal value, the compiler replaces the variable reference with the literal
itself.

Similarly to DSE, executing this replacement alters variable usage throughout
the graph, requiring data flow maps to be recalculated before the next pass.

\subsubsection{The optimization loop}

These three passes work side by side to provide an ultimately optimized graph.
For example, applying Constant Propagation might expose new literal values that
can be squashed together via Constant Folding. This folded constant might then
cause a previously useful variable to become obsolete, triggering Dead Store
Elimination.

To maximize efficiency, the main \Code{optimize()} function orchestrates these
passes using a fixpoint iteration loop. It continuously cycles through
\Code{doDSE()}, \Code{doConstFold()} and \Code{doConstProp()} in sequence. The
loop only terminates when a full cycle completes without any of the three passes
reporting a change to the graph or the AST, guaranteeing that the code is as
optimized as possible before code generation.

\subsubsection{Graph visualization}

The compiler supports generating a DOT graph after optimizations have been
applied. To do so, the following command is available:
\begin{verbatim}
$ ./run.sh miniimp -f code.mi -g -o
\end{verbatim}

It is possible to use the \texttt{-o} flag together with analysis-specific graph
generation flags, like \texttt{-ad}.

\subsection{Code generation}

The final phase of the compiler is to translate the verified and optimized
control flow graph into an intermediate representation. The choice of the
project is to rely on LLVM for it.

The entry point for this is the \Code{codegen()} function. Before emitting any
code, the compiler guarantees correctness by running the validation and
optimization passes, and then uses the \Code{maximizeGraph()} helper to group
the minimal nodes into larger basic blocks, which is ideal for LLVM IR
generation as seen previously.

\subsubsection{Variable allocation and the \texttt{alloca} trick}

A difficult problem when generating LLVM IR directly is adhering to its strict
\textit{Static Single Assignment} form, which requires the complex calculation
of \textit{phi} nodes for variables that change values across different control
flow branches. While the maximal graph may come in handy for this specific type
of problem, I ultimately decided to bypass this and implement what is commonly
known as the \texttt{alloca} trick, as advised in the slides. This is used for
example by the Rust compiler as well.

By invoking the \Code{extractAllVars()} helper, the compiler identifies every
unique identifier used throughout the program, including input, outputs and
intermediate variables. At the very beginning of the LLVM function definition, a
dedicated memory slot of the stack is allocated for each of these variables
using the \texttt{alloca} instruction.

This architecture allows the rest of the generated code to safely update
variables using standard \texttt{load} and \texttt{store} operations, deferring
the SSA conversion to LLVM's external \texttt{mem2reg} optimization pass.

\subsubsection{Expression translation}

To translate individual operations into machine-level instructions, the compiler
relies on two recursive helper functions: \Code{genNumExpr()} for aritmetic and
\Code{genBoolExpr()} for boolean logic.

These functions behave very similarly to the interpreter found in the
\path{engine.ts} module, traversing the expression trees bottom-up. For every
sub-expression evaluated, the compiler allocates a new, sequentially numbered
temporary LLVM register (e.g., \texttt{\%1}, \texttt{\%2}). Variables are loaded
from their stack pointers into these temporary registers, and mathematical
operations are mapped directly to their LLVM equivalent (such as \texttt{add},
\texttt{sub}, or \texttt{mul}).

For boolean logic, the compiler utilizes \texttt{and}, translates the
\texttt{not} operator into a \texttt{xor} instruction against the value 1, and
evaluates less-than comparisons using the \texttt{icmp slt} opcode.

\subsubsection{Graph traversal and block emission}

The final structure of the LLVM IR is constructed using a depth-first search
algorithm over the maximal graph, implemented via the \Code{traverse()} inner
function. As the routine visits each block, it assigns a unique label (such as
\texttt{block\_0}) to define a proper LLVM basic block.

Within each block, sequential assignment commands are evaluated, and their
resulting temporary registers are stored directly into stack memory. For
conditional blocks, the resulting condition value is stored into a register,
which is then used in the conditional \texttt{br} instruction.

In LLVM, every basic block must strictly terminate with a branching or returning
instruction. The \texttt{traverse()} crawler handles this dynamically based on
the block's type:
\begin{itemize}
    \item \textbf{Conditional blocks} evaluate their boolean condition and emit
        a conditional branch instruction (\texttt{br i1}) tat jumps to either
        the true or false block labels.
    \item \textbf{Standard blocks} that have a subsequent step emit an
        unconditional branch (\texttt{br}) to their single successor node.
    \item \textbf{Exit blocks} load the final computed value from the program's
        designated output variable and emit a \texttt{ret} instruction to return
        that integer to the caller.
\end{itemize}

\end{document}
